---
title: "tapSquares Decision Modeling"
subtitle: "Parsing session logs, allowing for heterogeneous players, and exporting an agent-behavior CSV for the ABM interpreter"
output:
  html_document:
    toc: true
    toc_float: true
    df_print: paged
---

<!--
HOW TO USE THIS FILE (R Markdown basics)
=============================================================
This is an R Markdown notebook (.Rmd) -- plain-text explanation (like
this comment) mixed with "chunks" of R code, delimited by ```{r} ... ```.
Run a chunk with the green play arrow at its top-right or Cmd/Ctrl+Enter
with your cursor in it; run top to bottom the first time through, since
later chunks depend on objects earlier ones create. "Knit" runs the
whole thing and produces one HTML report.

PIPELINE_CORE v1 -- reconciled from three copies that had diverged
across games (CropRaider/PES were untouched from the original;
Lantana had added an `excluded_predictors` mechanism and hardened
round/phase into a permanent exclusion; Tiger only differed in a
commented-out toggle). This version is the merge point going forward
-- see CHANGELOG.md in this folder for the full list of what changed
and why. Game-specific behavior belongs in that game's
`variableRoles.csv` (optional, see Step 4a below), never in an edited
copy of this file.

What changed in this version, compared to the one that ran clean
end-to-end on real CropRaider data:
  1. Round/phase (or year, or whatever a game calls its counter) are
     no longer unconditionally barred from being predictors. Some
     games embed real structure in that counter (a multi-year climate
     cycle, a seasonal effect) that a player could plausibly learn
     even with no explicit variable telling them about it. Step 4a
     reads an optional `variableRoles.csv` that can opt a counter
     variable in, and choose how it's encoded (raw numeric, cyclical
     sin/cos for a known period, or categorical) so the model can
     extrapolate to round/phase values it never saw in training
     instead of erroring or silently ignoring a real effect. Default
     behavior with no config file present, or a counter not listed in
     it, is unchanged: excluded, exactly as before.
  2. Land-ownership variables (`ownedBy`, `village`, and similar) get
     explicit handling: a configured sentinel value (e.g. 0) is kept
     as its own real "unassigned" factor level rather than being
     treated as just another number on a continuous scale. This
     matters because ownership in tapSquares games is sometimes
     player-selected, sometimes randomly/game-assigned, and some
     patches can stay unassigned the whole game -- the model needs to
     tell those cases apart, not average over them.
  3. `excluded_predictors` (from the Lantana branch) is kept as-is: a
     simple named-variable dropout list for anything that's a real
     predictor you just don't want in this particular fit.
  4. At the end, if `sessionList.csv` is present in the working
     directory, this notebook now writes `sessionList_ABM.csv` itself
     -- copying every column over and adding the three that point at
     this notebook's own outputs (`agent_behavior_weights.csv`,
     `agent_behavior_reference.csv`, `outcomePlotList.csv`). That
     wiring used to be a manual, error-prone edit; now it's automatic.
     Any other tuning for the ABM run (agent counts, world size, etc.)
     is still a manual edit on top of the generated file.

Honesty check, carried forward: steps 1-3 (parsing, pivoting, the
original single dedup+join+pooled-model path) have run against real
data and work. Steps 7-9 (per-player fits, clustering, per-class
refit, long-format export) were carefully traced by hand but are less
battle-tested -- treat a first error there as a real bug, not caution.
Steps 4a/4b (variable-role handling) are new in this version and have
only been checked against synthetic data, not real session logs yet.
-->

## Setup

```{r setup, message = FALSE}
# knitr/rmarkdown must be installed BEFORE this file can even be
# opened/knit (RStudio's rendering pipeline depends on them). If you see
# "there is no package called 'knitr'", run in the Console first:
#   install.packages(c("rmarkdown", "knitr"))

needed <- c("rmarkdown", "knitr", "readr", "dplyr", "stringr", "tidyr", "purrr", "nnet", "ggplot2")
missing_pkgs <- needed[!needed %in% installed.packages()[, "Package"]]
if (length(missing_pkgs) > 0) install.packages(missing_pkgs)

knitr::opts_chunk$set(warning = TRUE, message = FALSE)

library(readr)
library(dplyr)
library(stringr)
library(tidyr)
library(purrr)
library(nnet)
library(ggplot2)
```

## Point this at your session log(s)

```{r path}
# Default: every .csv in sample_sessions/. Swap to a single file (or a
# short explicit vector) once you're working with one real session:
#   log_paths <- "real_session.csv"
log_paths <- list.files("sample_sessions", pattern = "\\.csv$", full.names = TRUE)

# MANIFEST.csv (and anything else non-log dropped in this folder, e.g. a
# README) isn't a session log -- exclude by name rather than assuming the
# folder only ever holds log files.
log_paths <- log_paths[!grepl("MANIFEST", log_paths, ignore.case = TRUE)]

if (length(log_paths) == 0) {
  stop("log_paths is empty -- check your working directory (", getwd(), ").")
}
missing_files <- log_paths[!file.exists(log_paths)]
if (length(missing_files) > 0) {
  stop("Can't find: ", paste(missing_files, collapse = ", "), " (working directory: ", getwd(), "). ",
       "Check the path, and Session > Set Working Directory in RStudio if needed.")
}

log_paths
```

## Step 1: parse one file into tagged tables

```{r parse-function}
tag_schema <- list(
  STATE             = c("round", "phase", "x", "y", "name", "value"),
  CHOICE            = c("round", "phase", "player", "x", "y", "actionID", "timestamp"),
  CONFIRM           = c("round", "phase", "player", "timestamp"),
  VILLAGE           = c("round", "phase", "player", "x", "y", "status", "timestamp"),
  VILLAGECONFIRMED  = c("village", "x", "y", "status", "timestamp"),
  PRIVATE           = c("round", "phase", "player", "x", "y", "status", "timestamp"),
  PRIVATECONFIRMED  = c("player", "x", "y", "status", "timestamp"),
  SCORE             = c("round", "phase", "player", "currentScore", "totalScore", "resources"),
  PHASE             = c("label", "timestamp")
)

build_table <- function(split_lines, tags_found, tag, session_id) {
  cols <- tag_schema[[tag]]
  rows <- split_lines[tags_found == tag]
  rows <- map(rows, ~ .x[-1])
  good <- map_lgl(rows, ~ length(.x) == length(cols))
  if (any(!good)) {
    warning(sprintf("%s (%s): dropped %d malformed line(s) with the wrong field count",
                     tag, session_id, sum(!good)))
  }
  rows <- rows[good]
  if (length(rows) == 0) {
    empty <- setNames(as.list(rep(character(0), length(cols))), cols)
    df <- tibble::as_tibble(empty)
  } else {
    df <- as.data.frame(do.call(rbind, rows), stringsAsFactors = FALSE)
    names(df) <- cols
    df <- tibble::as_tibble(df)
  }
  df$session_id <- session_id
  df
}

parse_one_log <- function(path) {
  session_id <- tools::file_path_sans_ext(basename(path))
  raw_lines <- read_lines(path)
  split_lines <- str_split(raw_lines, fixed("|"))
  tags_found  <- map_chr(split_lines, ~ .x[1])

  list(
    state   = build_table(split_lines, tags_found, "STATE", session_id),
    choice  = build_table(split_lines, tags_found, "CHOICE", session_id),
    confirm = build_table(split_lines, tags_found, "CONFIRM", session_id),
    score   = build_table(split_lines, tags_found, "SCORE", session_id)
  )
}
```

## Step 2: parse every file and combine

```{r parse-all}
parsed_sessions <- map(log_paths, parse_one_log)

state_long <- map(parsed_sessions, "state")   %>% bind_rows()
choice     <- map(parsed_sessions, "choice")  %>% bind_rows()
confirm    <- map(parsed_sessions, "confirm") %>% bind_rows()
score      <- map(parsed_sessions, "score")   %>% bind_rows()

to_num <- function(df, cols) df %>% mutate(across(all_of(cols), as.numeric))

state_long <- to_num(state_long, c("round", "phase", "x", "y"))
choice     <- to_num(choice, c("round", "phase", "player", "x", "y", "actionID"))
score      <- to_num(score, c("round", "phase", "player", "currentScore", "totalScore", "resources"))

cat("sessions:", n_distinct(choice$session_id), "\n")
cat("state rows:", nrow(state_long), " choice rows:", nrow(choice),
    " confirm rows:", nrow(confirm), " score rows:", nrow(score), "\n")

choice %>% group_by(session_id) %>% summarise(n_players = n_distinct(player))
```

## Step 3: reshape state from long to wide

```{r pivot-state}
state_wide <- state_long %>%
  pivot_wider(id_cols = c(session_id, round, phase, x, y), names_from = name, values_from = value)

smart_coerce <- function(x) {
  if (all(x %in% c("true", "false", NA))) return(x == "true")
  numeric_try <- suppressWarnings(as.numeric(x))
  if (!any(is.na(numeric_try) & !is.na(x))) return(numeric_try)
  x  # leave as character -- e.g. playerHarvestRound's "[1 2 3]" list literal
}

state_wide <- state_wide %>% mutate(across(-c(session_id, round, phase, x, y), smart_coerce))

list_shaped <- names(state_wide)[map_lgl(state_wide, ~ is.character(.x) && any(str_detect(.x, fixed("[")), na.rm = TRUE))]
if (length(list_shaped) > 0) {
  message("List-shaped columns left as text, not used as predictors below: ",
          paste(list_shaped, collapse = ", "))
}
```

## Step 4: collapse taps into real decisions, without losing the taps themselves

One row per action, per the earlier reasoning (a turn can span several
patches, so one-row-per-turn doesn't fit a rectangular dataset). But
`update-current-choice` cycles through options each time a player taps
the *same* patch again, so only the last tap per patch is the real
decision -- the earlier ones are menu-browsing, not separate choices.

The count is kept as `n_taps_patch` (how many times this specific
patch was retapped) and `n_taps_turn` (how many total taps happened
across every patch this player touched that round/phase). Read these
two differently: `n_taps_turn` is a reasonable candidate *predictor*
-- it describes how much deliberation a player brought to the whole
turn, independent of which specific action they ended up taking
anywhere in it. `n_taps_patch` is murkier as a predictor of the action
chosen *on that same patch*, since the cycling order itself is fixed
by the game code. Left in as a candidate either way -- just worth
treating with more suspicion than the state variables.

```{r dedup-choice}
cat("choice rows before collapsing cycling taps:", nrow(choice), "\n")

choice <- choice %>%
  group_by(session_id, round, phase, player, x, y) %>%
  mutate(n_taps_patch = n()) %>%
  ungroup() %>%
  group_by(session_id, round, phase, player) %>%
  mutate(n_taps_turn = n()) %>%
  ungroup()

choice_real <- choice %>%
  group_by(session_id, round, phase, player, x, y) %>%
  slice_tail(n = 1) %>%
  ungroup()

cat("choice rows after collapsing to real decisions:", nrow(choice_real), "\n")

# tapSquares uses -99 as the sentinel for "explicitly chose nothing here."
# TRUE keeps it as a real outcome level (modeling "did they act at all,
# and on what" together); FALSE models only the choice of *which* action,
# conditional on having taken one. Either is defensible -- flip this and
# re-run rather than picking one silently.
include_no_action_choice <- TRUE
if (!include_no_action_choice) {
  choice_real <- choice_real %>% filter(actionID != -99)
  cat("choice rows after dropping 'no action' (-99):", nrow(choice_real), "\n")
}
```

## Step 4a: load this game's variable-roles config (optional)

`variableRoles.csv` is per-game, optional, and lives alongside the
session logs. It tells this notebook how to treat three kinds of
variable that need more than "is it numeric or logical":

- **counters** (`round`, `phase`, `year`, or whatever a game calls its
  turn/cycle number). Default (not listed, or no config file at all)
  is `exclude` -- identical to every prior version of this notebook.
  List one here with `encoding = numeric` to test it as a plain linear
  predictor, or `encoding = cyclical` with a `period` to test it as a
  repeating cycle (encoded as sin/cos pairs so the model can
  extrapolate to a round number it never saw in training, rather than
  needing every value seen during fitting).
- **ownership** (`ownedBy`, `village`, or similar). `role =
  ownership_individual`/`ownership_group` treats the variable as a
  category (its sentinel value, e.g. 0 for unassigned, becomes its own
  distinct level) rather than a number the model averages over.
  `unassignedValue` is documentation for humans reading the config --
  not consumed by this notebook -- so the raw value stays the level
  text, matching what the ABM generator reads back from the patch.
- **treatment candidates**: `treatmentCandidate = TRUE` doesn't change
  modeling at all -- it's a flag carried through to the export/notes
  so an ABM user knows which variables are the ones worth sweeping.

A blank template with all of this documented lives at
`variableRoles_TEMPLATE.csv` in pipeline_core.

```{r load-variable-roles}
role_schema <- c("variable", "role", "encoding", "period", "unassignedValue", "treatmentCandidate", "notes")

if (file.exists("variableRoles.csv")) {
  variable_roles <- read_csv("variableRoles.csv", col_types = cols(.default = "c"), show_col_types = FALSE)
  missing_cols <- setdiff(role_schema, names(variable_roles))
  if (length(missing_cols) > 0) {
    stop("variableRoles.csv is missing column(s): ", paste(missing_cols, collapse = ", "))
  }
  cat("Loaded variableRoles.csv:", nrow(variable_roles), "variable(s) configured.\n")
} else {
  variable_roles <- tibble(variable = character(), role = character(), encoding = character(),
                            period = character(), unassignedValue = character(),
                            treatmentCandidate = character(), notes = character())
  message("No variableRoles.csv found -- using defaults (round/phase excluded, everything else auto-detected).")
}

variable_roles
```

## Step 4b: apply counter and ownership encodings to the wide state table

```{r apply-variable-roles}
counter_defaults <- c("round", "phase")  # always excluded unless explicitly overridden below

# Cyclical counters: replace the raw column with sin/cos pairs.
cyclical_rows <- variable_roles %>% filter(role == "counter", encoding == "cyclical")
for (i in seq_len(nrow(cyclical_rows))) {
  v <- cyclical_rows$variable[i]
  p <- suppressWarnings(as.numeric(cyclical_rows$period[i]))
  if (v %in% names(state_wide) && !is.na(p) && p > 0) {
    state_wide[[paste0(v, "_sin")]] <- sin(2 * pi * state_wide[[v]] / p)
    state_wide[[paste0(v, "_cos")]] <- cos(2 * pi * state_wide[[v]] / p)
    message(v, ": encoded as cyclical (period = ", p, ") -> ", v, "_sin, ", v, "_cos")
  } else if (v %in% names(state_wide)) {
    warning(v, ": role=counter, encoding=cyclical but no valid 'period' given -- left out of the model.")
  }
}

# Numeric counters: leave the raw column in place, just don't structurally exclude it below.
numeric_counter_vars <- variable_roles %>% filter(role == "counter", encoding == "numeric") %>% pull(variable)

# Categorical counters (rare -- won't extrapolate to an unseen round/phase value,
# only use this if you specifically want "round 3" treated as a distinct category).
categorical_counter_vars <- variable_roles %>% filter(role == "counter", encoding == "categorical") %>% pull(variable)
for (v in categorical_counter_vars) {
  if (v %in% names(state_wide)) state_wide[[v]] <- factor(state_wide[[v]])
}

# Ownership variables: treated as a category (factor), not a number on a
# continuous scale, so the sentinel (e.g. 0 = unassigned) is just another
# distinct level rather than something the model averages toward. The raw
# value is kept AS the level text (e.g. "0", "1", "2") rather than
# relabeled to a word like "unassigned" -- R's dummy-coded coefficient
# name is then <varname><rawvalue> (e.g. "village0"), which the ABM
# generator's get-feature-value resolves by reading that same patch
# variable's current value back as text. Relabeling to a word here would
# require the ABM side to separately know the unassignedValue mapping
# from variableRoles.csv just to reverse it -- unnecessary coupling for
# no modeling benefit, since factor() already gives 0 its own level
# either way. unassignedValue in variableRoles.csv is documentation for
# humans reading the config, not consumed here.
ownership_rows <- variable_roles %>% filter(role %in% c("ownership_individual", "ownership_group"))
for (i in seq_len(nrow(ownership_rows))) {
  v <- ownership_rows$variable[i]
  if (v %in% names(state_wide)) {
    state_wide[[v]] <- factor(as.character(state_wide[[v]]))
    message(v, ": encoded as ownership factor (levels: ",
            paste(levels(state_wide[[v]]), collapse = ", "), ")")
  }
}

# round/phase stay excluded from the id-column list below UNLESS they were
# given a non-"exclude" encoding above (numeric, cyclical, or categorical).
counter_included <- variable_roles %>%
  filter(role == "counter", encoding %in% c("numeric", "cyclical", "categorical")) %>%
  pull(variable)
counters_to_exclude <- setdiff(counter_defaults, counter_included)

cat("Counters excluded from modeling:", paste(counters_to_exclude, collapse = ", "), "\n")
if (length(counter_included) > 0) cat("Counters included (see encoding above):", paste(counter_included, collapse = ", "), "\n")
```

## Step 5: build one row per decision

```{r model-data}
model_data <- choice_real %>%
  left_join(state_wide, by = c("session_id", "round", "phase", "x", "y")) %>%
  left_join(score, by = c("session_id", "round", "phase", "player"))

# Always-bookkeeping columns: join keys / identifiers, never candidate
# predictors regardless of configuration. round/phase are handled
# separately just above -- excluded by default, includable via
# variableRoles.csv -- since unlike these, a game *can* meaningfully
# encode real state in its counter.
id_cols <- c("session_id", "player", "x", "y", "actionID", "timestamp")
id_cols <- union(id_cols, counters_to_exclude)

predictor_cols <- setdiff(names(model_data), id_cols)
predictor_cols <- predictor_cols[map_lgl(model_data[predictor_cols],
                                          ~ is.numeric(.x) || is.logical(.x) || is.factor(.x))]

## Choose which real predictors the choice model may use
##
## `session_id`, `player`, `x`, `y`, `actionID`, and `timestamp` are
## never candidates -- they're bookkeeping from the log format itself
## (join keys and identifiers), not variables that exist anywhere in
## the actual game/ABM. round/phase are excluded by the same logic
## UNLESS variableRoles.csv opted them in (Step 4a/4b) -- see that
## section for why a game might want to.
##
## What *does* belong in excluded_predictors below: any predictor
## that's real (a patches-own variable that's actually in this game,
## or a score-derived feature like `resources`) but that you know
## isn't meaningful for this analysis and want dropped without
## touching the data or the model-fitting code. Empty by default.
excluded_predictors <- c()

predictor_cols <- setdiff(predictor_cols, excluded_predictors)

cat("Excluded from the choice model by name (edit excluded_predictors above to add any):\n",
    if (length(excluded_predictors) == 0) "(none)" else paste(excluded_predictors, collapse = ", "), "\n")

# tapSquares names boolean patches-own variables with a trailing "?"
# (water?, cropped?, ...) -- not valid bare terms in an R formula, so
# every predictor gets backtick-wrapped regardless of whether it needs
# it. Works the same for factor columns (ownership, categorical
# counters) as for numeric/logical ones.
predictor_terms <- paste0("`", predictor_cols, "`")

cat("Predictors used (includes n_taps_patch, n_taps_turn):\n", paste(predictor_cols, collapse = ", "), "\n")

# Coefficients from separately-fit models are only comparable if every
# model uses the same reference/baseline outcome category. Fix one
# explicitly and reuse it for every fit below (pooled, per-player,
# per-class) rather than letting each fit pick its own default.
#
# The reference category MUST be -99 ("do nothing") whenever it's present
# in the data, not just "whichever action happens to be most common."
# A multinomial logit's reference level always scores exactly 0 -- no
# fitted coefficients, no dependence on state at all, by construction --
# and the ABM (choose-actions-for-agent, tapSquares_ABM_generator.nlogox)
# picks every action by comparing its fitted score against -99's fitted
# score (the margin over "doing nothing"). That comparison is only
# mathematically sound when -99 itself is the reference: then every real
# action's score directly IS its own log-odds relative to doing nothing,
# and "margin > 0" means exactly what it's supposed to. Pick anything
# else as reference and that ONE action instead gets pinned to a flat,
# state-independent 0 score forever, while -99 gets real (often
# comparatively favorable) fitted coefficients -- so the reference action
# ends up silently, permanently excluded from every ABM agent's choices,
# for a reason that has nothing to do with real behavior, just an
# artifact of which action happened to get picked as the baseline.
# Confirmed live: Crouching Tiger's synthetic training data (near-uniform
# random action choice, no real signal) happened to make actionID 4
# (buildFence) the most-frequent action by sampling noise alone -- it
# became the reference, and fence-building was never chosen by any ABM
# agent afterward, regardless of session/round/patch state.
if (-99 %in% model_data$actionID) {
  reference_action <- -99
} else {
  # -99 never appears in this session's data (every player acted every
  # turn) -- fall back to the most frequent real action, same as before.
  reference_action <- model_data %>% count(actionID, sort = TRUE) %>% slice(1) %>% pull(actionID)
  message("Note: -99 (\"do nothing\") never appears in this session's data -- ",
          "falling back to the most frequent action (", reference_action,
          ") as the reference category instead.")
}
model_data <- model_data %>% mutate(actionID_f = relevel(factor(actionID), ref = as.character(reference_action)))
cat("Reference category for all model fits below: actionID =", reference_action, "\n")

form <- as.formula(paste("actionID_f ~", paste(predictor_terms, collapse = " + ")))
can_fit_model <- n_distinct(model_data$actionID) >= 2
if (!can_fit_model) {
  message("Only one action type appears in this data -- can't fit a choice model yet.")
}

model_data
```

## Step 6: a single pooled model, as a baseline

Useful as a sanity check and a point of comparison against the
per-player/per-class versions below -- if the pooled and per-class
models end up looking nearly identical, that's itself informative (not
much heterogeneity to find, at least not yet, with this much data).

```{r fit-pooled}
if (can_fit_model) {
  fit <- multinom(form, data = model_data, trace = FALSE)
  summary(fit)
}
```

```{r plot-coefs}
if (exists("fit")) {
  coefs <- as.data.frame(coef(fit))
  coefs$outcome_level <- rownames(coefs)
  coefs_long <- coefs %>% pivot_longer(-outcome_level, names_to = "term", values_to = "weight")

  ggplot(coefs_long, aes(x = term, y = weight, fill = outcome_level)) +
    geom_col(position = "dodge") +
    coord_flip() +
    labs(title = "Pooled model: fitted feature weights per action",
         x = NULL, y = "coefficient (log-odds scale, relative to the reference action)") +
    theme_minimal()
}
```

## Step 7: let players differ -- fit one model per player, then cluster

This is a practical two-step *approximation* to a true latent-class
discrete choice model, not the real thing. A proper joint estimator
(e.g. the `gmnl` package's latent-class multinomial logit) fits all
classes simultaneously via EM, sharing information across players in a
statistically principled way. That needs a data format (`mlogit`-style
long-by-alternative, via `dfidx`) that's easy to get subtly wrong
without being able to test it live, and with only a few players from
one session there often isn't yet enough data to make a joint
estimator worth the risk. This fit-per-player-then-cluster version
reuses only the already-validated `multinom()` machinery, so it's far
more likely to actually run -- worth upgrading to `gmnl` once there
are enough sessions/players for a joint estimate to be meaningfully
more powerful than this approximation.

**With only a handful of players, treat this whole step as a
mechanics check, not a real result** until there's enough data to say
otherwise.

```{r fit-per-player}
fit_one_player <- function(pid) {
  player_data <- model_data %>% filter(player == pid)
  if (n_distinct(player_data$actionID) < 2) {
    message("Player ", pid, ": only one action type observed -- skipping.")
    return(NULL)
  }
  tryCatch(
    multinom(form, data = player_data, trace = FALSE),
    error = function(e) {
      message("Player ", pid, ": model failed to fit (", conditionMessage(e), ")")
      NULL
    }
  )
}

if (can_fit_model) {
  player_ids <- sort(unique(model_data$player))
  player_fits <- set_names(map(player_ids, fit_one_player), player_ids)
  player_fits <- compact(player_fits)
  cat("Per-player models fit successfully for players:", paste(names(player_fits), collapse = ", "), "\n")
}
```

```{r cluster-players}
# nnet::multinom's coef() returns a matrix when there are 3+ outcome
# levels, but a plain vector when there are exactly 2 -- a real
# possibility with this little data per player. Rather than guess at
# reconstructing the matrix shape without being able to test it, players
# who land in that case get skipped here with a message, not silently
# mishandled.
extract_coefs_long <- function(model, player_id) {
  cm <- coef(model)
  if (is.null(dim(cm))) {
    message("Player ", player_id, ": exactly two action levels in this player's ",
            "data, which changes coef()'s output shape -- skipping in clustering.")
    return(NULL)
  }
  as.data.frame(cm) %>%
    tibble::as_tibble() %>%
    mutate(outcome_level = rownames(cm), player = player_id) %>%
    pivot_longer(-c(outcome_level, player), names_to = "feature", values_to = "weight")
}

player_classes <- NULL

if (can_fit_model && length(player_fits) > 0) {
  player_coefs_long <- map2(player_fits, names(player_fits), extract_coefs_long) %>%
    compact() %>%
    bind_rows()

  if (nrow(player_coefs_long) > 0) {
    player_coefs_wide <- player_coefs_long %>%
      mutate(col_name = paste(outcome_level, feature, sep = "__")) %>%
      select(player, col_name, weight) %>%
      pivot_wider(names_from = col_name, values_from = weight)

    # a player who never took a given action won't have a coefficient for
    # it -- treating that as 0 for clustering purposes only (not a real
    # statistical imputation) is a simplification worth revisiting once
    # more players make full coverage more likely.
    player_coefs_matrix <- player_coefs_wide %>%
      select(-player) %>%
      mutate(across(everything(), ~ replace(.x, is.na(.x), 0))) %>%
      as.matrix()
    rownames(player_coefs_matrix) <- player_coefs_wide$player

    n_players_clusterable <- nrow(player_coefs_matrix)
    col_sds <- apply(player_coefs_matrix, 2, sd)
    usable_cols <- col_sds > 1e-8

    if (n_players_clusterable < 3 || sum(usable_cols) == 0) {
      message("Not enough variation across players to cluster meaningfully yet ",
              "(", n_players_clusterable, " player(s), ", sum(usable_cols),
              " varying feature(s)) -- skipping clustering, will fall back to ",
              "the pooled model for export below.")
    } else {
      k <- min(2, n_players_clusterable - 1)
      set.seed(1)
      km <- kmeans(scale(player_coefs_matrix[, usable_cols, drop = FALSE]), centers = k)
      player_classes <- tibble(player = as.numeric(rownames(player_coefs_matrix)),
                                class = paste0("class_", km$cluster))
    }
  }
}

player_classes
```

## Step 8: refit one pooled model per discovered class

```{r fit-per-class}
class_fits <- list()

if (!is.null(player_classes)) {
  model_data_classed <- model_data %>% left_join(player_classes, by = "player")

  fit_one_class <- function(cid) {
    class_data <- model_data_classed %>% filter(class == cid)
    if (n_distinct(class_data$actionID) < 2) {
      message("Class ", cid, ": only one action type observed -- skipping.")
      return(NULL)
    }
    tryCatch(
      multinom(form, data = class_data, trace = FALSE),
      error = function(e) {
        message("Class ", cid, ": failed to fit (", conditionMessage(e), ")")
        NULL
      }
    )
  }

  class_ids <- sort(unique(model_data_classed$class))
  class_fits <- set_names(map(class_ids, fit_one_class), class_ids)
  class_fits <- compact(class_fits)
  cat("Per-class models fit successfully for:", paste(names(class_fits), collapse = ", "), "\n")
}
```

## Step 9: export the agent-behavior CSV, and the ABM session list, for the interpreter

Long format: `agent_type` (which class of player this row describes),
`decision_stage` (fixed at `"action_choice"` for everyone right now --
a placeholder column, not yet doing anything, since nothing in the
data so far has justified a genuine two-stage strategy-then-tactic
model), `outcome_option` (which action), `feature`, and `weight`.
Falls back to the single pooled model, tagged `agent_type = "pooled"`,
if clustering didn't produce usable per-class fits.

```{r export}
extract_export_rows <- function(model, agent_type) {
  cm <- coef(model)
  if (is.null(dim(cm))) {
    message(agent_type, ": exactly two action levels -- skipping in export (see Step 7's note).")
    return(NULL)
  }
  as.data.frame(cm) %>%
    tibble::as_tibble() %>%
    mutate(outcome_option = rownames(cm)) %>%
    pivot_longer(-outcome_option, names_to = "feature", values_to = "weight") %>%
    mutate(agent_type = agent_type, decision_stage = "action_choice") %>%
    select(agent_type, decision_stage, outcome_option, feature, weight)
}

export_rows <- tibble()

if (length(class_fits) > 0) {
  export_rows <- map2(class_fits, names(class_fits), extract_export_rows) %>% compact() %>% bind_rows()
}

if (nrow(export_rows) == 0 && exists("fit")) {
  message("No usable per-class models -- exporting the pooled model instead, as agent_type = 'pooled'.")
  export_rows <- extract_export_rows(fit, "pooled")
}

if (nrow(export_rows) > 0) {
  write_csv(export_rows, "agent_behavior_weights.csv")

  # The ABM interpreter needs to know which actionID was the reference/
  # baseline category, since a multinomial logit's reference level has no
  # row in coef() at all -- its score is implicitly 0, everything else is
  # a log-odds relative to it. Writing it out explicitly here rather than
  # making the NetLogo side infer it.
  write_csv(tibble(reference_actionID = reference_action), "agent_behavior_reference.csv")
}

export_rows
```

```{r export-sessionlist-abm}
# Wire the ABM run's sessionList.csv automatically: same columns as the
# game's own sessionList.csv, plus the three pointers to this notebook's
# own outputs. Any other ABM-specific tuning (agent counts, a bigger
# world, etc.) is still a manual edit on top of the file this writes --
# only the error-prone pointer wiring is automated here.
if (file.exists("sessionList.csv") && nrow(export_rows) > 0) {
  session_list <- read_csv("sessionList.csv", show_col_types = FALSE)
  session_list$behaviorListFileName     <- "agent_behavior_weights.csv"
  session_list$referenceActionFileName  <- "agent_behavior_reference.csv"
  session_list$outcomePlotListFileName  <- "outcomePlotList.csv"
  # na = "" (not readr's default "NA") -- sessionList.csv can legitimately
  # leave an optional column blank (e.g. patch_color_property, meaning "no
  # property-driven shading, use the default color"). read_csv() reads
  # that blank cell as R's NA; write_csv()'s default na="NA" would then
  # serialize it back out as the literal 3-character string "NA", not an
  # empty cell. The ABM generator's generic CSV-driven config loader has
  # no way to tell "the text NA" apart from a real value, so it tries to
  # runresult "NA" as a NetLogo expression and crashes ("Nothing named NA
  # has been defined") the moment that column is used -- confirmed live,
  # Cut the Creep's patch_color_property. na="" keeps a blank cell blank
  # all the way through the round-trip, matching what a human-authored
  # sessionList.csv would have.
  write_csv(session_list, "sessionList_ABM.csv", na = "")
  cat("Wrote sessionList_ABM.csv (", nrow(session_list), "row(s), from sessionList.csv plus the three output pointers).\n")
} else if (!file.exists("sessionList.csv")) {
  message("No sessionList.csv in the working directory -- skipped writing sessionList_ABM.csv. ",
          "Copy the game's sessionList.csv here (or point setwd() at the game folder) and re-run this chunk.")
}
```

## Notes / what this notebook deliberately doesn't do yet

- **Own-patch state only.** Predictors come only from the patch a
  player acted on, not the fuller board-wide / multi-scale feature set
  discussed earlier.
- **No visibility-treatment handling** yet.
- **Player identity isn't tracked across sessions** -- `player` numbers
  reset per session.
- **Assumes one consistent game per batch of files** (same
  `patchPropertyList` schema across sessions).
- **The per-player-then-cluster approach is an approximation.** A real
  latent-class discrete choice model (`gmnl`, `model = "lc"`) is the
  more rigorous version, worth moving to once there's enough data
  across enough sessions/players to make joint EM estimation pay off
  over this simpler two-step version.
- **The two-stage `decision_stage` column is a placeholder.** Nothing
  currently populates more than one stage.
- **Exactly two available actions is a rough edge** for `coef()`'s
  output shape, handled by skipping (with a message) rather than
  guessing at a fix, in the per-player, clustering, and export steps.
- **Categorical counter/ownership levels unseen at prediction time**
  are this notebook's problem only insofar as they can't be fit on
  data that doesn't contain them. Making the *ABM generator* degrade
  gracefully (rather than erroring) when it encounters an unseen
  categorical level at runtime is separate NetLogo-side work, tracked
  alongside this file, not solved here.
